---
title: "00 - Data Exploration"
subtitle: "Tadpole gut microbiome: multilevel community assembly"
author: "Javi Rudolph"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    theme: flatly
    embed-resources: true
execute:
  warning: false
  message: false
---

## Overview

Initial exploration of the tadpole gut microbiome dataset from Correa et al. (2020), 
which investigates multilevel community assembly across three biogeographic areas 
(Brazil, Central Texas, East Texas). The core question is whether **pond of origin** 
(regional species pool) or **host species identity** (local filter) is more important 
in structuring the gut microbiome of tadpoles.

This notebook loads all raw data files, inspects their structure, and sets up a 
`phyloseq` object for downstream analysis.

---

## Setup

```{r}
#| label: libraries

library(tidyverse)
library(readxl)
library(janitor)
library(skimr)
library(phyloseq)
```

---

## Load raw data

### OTU table

Rows are samples, columns are ASVs (or vice versa — check orientation below).

```{r}
#| label: load-otu

otu_raw <- read_csv("../data/raw/otu_table.csv")
# Big files here, so careful

dim(otu_raw)
otu_raw[1:5, 1:8]          # peek at corner
colnames(otu_raw)[1:10]    # first few column names
```

Need to handle stuff with the big size:
```{r}
otu_mat <- otu_raw |> 
  column_to_rownames("SampleName") |> 
  as.matrix()

rm(otu_raw)   # free memory immediately
gc()          # trigger garbage collection
```

The OTU table contains 243 samples and 23,051 amplicon sequence variants (ASVs). Given the size of this matrix, we convert it to a base R matrix early and remove the raw tibble from memory to keep the session lean. Microbiome count matrices are typically very sparse so we avoid calling glimpse() or plotting directly on the raw object. Instead, we summarize first and plot summaries.

```{r}
# Sparsity - expect > 90% zeros
mean(otu_mat == 0)

# Reads per sample
sample_reads <- rowSums(otu_mat)
summary(sample_reads)

# ASV prevalence - how many samples each ASV appears in
asv_prevalence <- colSums(otu_mat > 0)
summary(asv_prevalence)
```

The ASV prevalence summary shows the extreme sparsity that is typical of microbiome count data. Of the 23,051 ASVs detected, the median prevalence is 1, meaning more than half of all ASVs appear in only a single sample. The mean prevalence of about 3 samples per ASV and a maximum of 93 confirm that while a small number of ASVs are broadly distributed, the vast majority are rare and pond-specific. This is consistent with Correa et al. (2020), who found that most bacteria occur in only a single pond. Because so many ASVs appear in only one sample and can't be distinguished from sequencing noise, we apply a minimum prevalence filter before moving on, which brings the dataset down to a more manageable size while keeping the ecologically relevant signal.

```{r}
# How many ASVs appear in only 1 sample?
sum(asv_prevalence == 1)
sum(asv_prevalence == 1) / ncol(otu_mat) * 100  # as percentage

# Visualize the prevalence distribution (summarize first!)
table(cut(asv_prevalence, breaks = c(0,1,2,5,10,50,100)))

otu_mat_filt <- otu_mat[, asv_prevalence > 1]
dim(otu_mat_filt)
```

Applying a minimum prevalence threshold of 2 samples removes 13,618 singleton ASVs, retaining 9,433 ASVs across all 243 samples. This represents about 41% of the original ASVs, with the removed taxa being those indistinguishable from sequencing artifacts.


### Taxonomy table

Maps each ASV to taxonomic ranks (Kingdom through Species based on Silva v128 
used in the original study).

```{r}
#| label: load-tax

tax_raw <- read_csv("../data/raw/tax_table.csv")

glimpse(tax_raw)
```

```{r}
tax_raw |> count(Kingdom)

tax_raw |> count(Phylum, sort = TRUE)

tax_raw |> 
  summarise(across(Kingdom:Genus, ~sum(is.na(.)) / n() * 100))
```


### Sample data

The sample metadata contains 243 samples across three sites, Boraceia (BR), DCNF (ET), and Ecolab (CT), with each site having sediment, water, and tadpole gut samples. Pond and species counts match the paper for Boraceia and DCNF, but Ecolab shows 13 ponds versus the 7 reported in Correa et al. (2020), which is worth investigating before downstream analysis. Ecolab also has notably fewer tadpole samples (26) relative to its pond count, so we'll check the site-specific metadata to understand what's going on there.

```{r}
#| label: load-sample-data

sample_meta <- read_csv("../data/raw/sample_data.csv")

glimpse(sample_meta)
```

```{r}
# Samples per site
sample_meta |> count(Site)

# Sample types per site
sample_meta |> count(Site, SampleType)

# Ponds per site
sample_meta |> distinct(Site, Pond) |> count(Site)

# Tadpole species per site
sample_meta |> 
  filter(SampleType == "tadpole") |> 
  distinct(Site, Species) |> 
  count(Site)
```



### EBB metadata

The Brazil metadata contains 93 samples across 6 ponds (Brejo, Castelo, Pedreira, R2, R3, and Bomba), with 81 tadpole gut samples and 6 water and sediment samples each, consistent with Correa et al. (2020). Tadpole rows include full taxonomic information at the family, genus, and species level, with Genus_Species being the most convenient identifier for downstream analysis. Coordinates are available for all samples, which will be useful for the dispersal limitation analysis. Environmental sample rows use pond name as a placeholder in the species columns rather than leaving them empty.

```{r}
#| label: load-ebb

ebb_meta <- read_csv("../data/raw/ebb_metadata.csv")

glimpse(ebb_meta)
```


```{r}
# See the tadpole rows where species info is actually populated
ebb_meta |> 
  filter(!Type %in% c("Substrate", "Water")) |> 
  select(SampleName, Site, family, frog_Genus, frog_Species, Genus_Species)

# How many sample types?
ebb_meta |> count(Type)

# All 6 Brazil ponds present?
ebb_meta |> distinct(Site)
```


### Ecolab metadata

The Ecolab metadata clarifies the apparent pond count discrepancy. Of the 13 ponds in the dataset, only 7 contained tadpoles (hastads == "y"), matching exactly the 7 Central Texas ponds reported in Correa et al. (2020). The remaining 6 ponds were sampled for water and sediment only and were not included in the published analysis. Four additional rows with missing pond information correspond to blank and failed samples that will be excluded going forward. The Ecolab metadata also contains additional covariates including tadpole size, source/sink classification, and coordinates for a subset of samples, which may be useful.

```{r}
#| label: load-ecolab

ecolab_meta <- read_csv("../data/raw/ecolab_metadata.csv")

glimpse(ecolab_meta)
```

```{r}
# How many NAs in pond?
ecolab_meta |> count(is.na(pond))

# What are those NA pond rows?
ecolab_meta |> filter(is.na(pond))

# Check hastads
ecolab_meta |> count(hastads, pond)
```


### DCNF metadata

Data from David Crockett and Angelina National Forests (East Texas, ET site).

```{r}
#| label: load-dcnf

dcnf_meta <- read_xlsx("../data/raw/DCNF_metadata.xlsx") |> 
  clean_names()

glimpse(dcnf_meta)
```

```{r}
# Fix numeric columns
dcnf_meta <- dcnf_meta |> 
  mutate(across(c(latitute, longitude, watertemp, conductivity, 
                  chlorophyll, do, p_h, gosner_stage), as.numeric))

# Check sample types
dcnf_meta |> count(type_of_sample)

# Check ponds
dcnf_meta |> distinct(pond)

# Explain the 132 vs 95 discrepancy
dcnf_meta |> 
  count(sample_name %in% sample_meta$SampleName)
```

```{r}
# Compare a few names directly
head(dcnf_meta$sample_name)
head(sample_meta |> filter(Site == "DCNF") |> pull(SampleName))

# Try case-insensitive match
dcnf_meta |> 
  count(toupper(sample_name) %in% toupper(sample_meta$SampleName))
```

Froggies only:
```{r}
dcnf_meta %>% filter(family %in% c("Ranidae", "Hylidae", "Microhylidae"))

# family being an environmental variable

# salmanders are:  Sirenidae, Ambystomatidae, and Salamandridae

# DCNF try with filtered singletons, 2... 

# 9 ponds
# we are trying to get the same analysis from figure 1 using JSDM
# don't lump froggie or not in the jsdm

dcnf_meta %>% 
  filter(family %in% c("Ranidae", "Sirenidae", "Ambystomatidae",
                       "Microhylidae", "Hylidae", "Salamandridae"
                       #, "WATER", "SUBSTRATE"
                       )) %>% 
  mutate(frog_or_not = ifelse(family %in% c("Ranidae", "Microhylidae", "Hylidae"),
                              "froggie", "not")) %>% 
  count(frog_or_not, pond) %>% 
  # count(family, pond) %>% 
  arrange(n)

dcnf_meta$pond %>% unique()
dcnf_meta %>% filter(pond == "87-1") #excluded from the paper

dcnf_meta %>% count(pond)
```




---

## Initial inspection

### Sample data summary

```{r}
#| label: sample-summary

skim(sample_meta)
```

How many samples per site and sample type?

```{r}
#| label: sample-counts

sample_meta |> 
  count(Site, SampleType)
```

How many tadpole species per pond?

```{r}

sample_meta |> 
  filter(SampleType == "tadpole") |>
  count(Pond, Site, Species)
```

How many water bodies (ponds) per site?

```{r}

sample_meta |> 
  count(Pond, Site)
```

### OTU table dimensions

```{r}
#| label: otu-dims
cat("OTU table dimensions (raw):     243 x 23051\n")
cat("OTU table dimensions (filtered):", nrow(otu_mat_filt), "x", ncol(otu_mat_filt), "\n")
```

### Taxonomy table summary
```{r}
#| label: tax-phylum-counts

tax_raw |> count(Phylum, sort = TRUE)
```

## Build phyloseq object
```{r}
#| label: build-phyloseq

# --- OTU matrix (use filtered) ---
# rows = samples, columns = ASVs
otu_mat_ps <- otu_mat_filt

# --- Taxonomy matrix ---
tax_mat <- tax_raw |>
  column_to_rownames(var = "SV") |>
  as.matrix()

# Keep only ASVs present in filtered OTU matrix
tax_mat <- tax_mat[colnames(otu_mat_ps), ]

# --- Sample data frame ---
samp_df <- sample_meta |>
  column_to_rownames(var = "SampleName")

# --- Assemble phyloseq object ---
ps <- phyloseq(
  otu_table(otu_mat_ps, taxa_are_rows = FALSE),
  tax_table(tax_mat),
  sample_data(samp_df)
)

ps
```




### Quick sanity checks

```{r}
#| label: phyloseq-checks

cat("Number of samples:", nsamples(ps), "\n")
cat("Number of ASVs:   ", ntaxa(ps), "\n")
cat("Sample variables: ", paste(sample_variables(ps), collapse = ", "), "\n")
cat("Taxonomic ranks:  ", paste(rank_names(ps), collapse = ", "), "\n")
```

Reads per sample distribution:

```{r}
#| label: reads-per-sample

reads_per_sample <- sample_sums(ps)

tibble(
  sample = names(reads_per_sample),
  n_reads = reads_per_sample
) |> 
  ggplot(aes(x = n_reads)) +
  geom_histogram(bins = 40, fill = "#3d6b8c", color = "white") +
  labs(
    title = "Distribution of reads per sample",
    x = "Number of reads",
    y = "Count"
  ) +
  theme_minimal()
```

---


## Session info

```{r}
#| label: session-info

sessionInfo()
```
