---
title: "01 - DCNF sjSDM"
subtitle: "Tadpole gut microbiome: Joint Species Distribution Model — East Texas (DCNF)"
author: "Javi Rudolph"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    theme: flatly
    embed-resources: true
execute:
  warning: false
  message: false
---

## Overview

This notebook focuses on the **David Crockett and Angelina National Forests (DCNF / East Texas)**
site, which has the richest sampling in the dataset: 9 ponds, 6 tadpole species, and the most
complete set of environmental covariates (dissolved oxygen, pH, conductivity, chlorophyll-a,
water temperature, and fish presence/absence).

The goal is to fit a **Joint Species Distribution Model (sjSDM)** to the tadpole gut microbiome
at DCNF, using pond-level environmental variables as predictors and latent factors to capture
residual co-occurrence structure among ASVs after accounting for measured environment.

---

## Setup

```{r}
#| label: libraries

library(tidyverse)
library(readxl)
library(janitor)
library(sjSDM)
```

---

## Load raw data

### OTU table

The full table has 243 samples × 23,051 ASVs across all three sites (Brazil, Central Texas,
DCNF). We load it, convert to a matrix immediately, and drop the tibble to save memory.

```{r}
#| label: load-otu

otu_raw <- read_csv("../data/raw/otu_table.csv")

otu_mat <- otu_raw |>
  column_to_rownames("SampleName") |>
  as.matrix()

rm(otu_raw)
gc()

```

Since this one is heavy, we want to apply the filters we explored in the previous notebook:

```{r}
# Sparsity summary
mean(otu_mat == 0)

sample_reads <- rowSums(otu_mat)
asv_prevalence <- colSums(otu_mat > 0)

summary(sample_reads)
summary(asv_prevalence)
table(cut(asv_prevalence, breaks = c(0, 1, 2, 5, 10, 50, 100, Inf)))
```
Choose what we use as the filter: 

```{r}
# --- change this threshold ---
min_prevalence <- 2

otu_mat_filt <- otu_mat[, asv_prevalence >= min_prevalence]
```

The output matrix is filtered, where rownames are the sample name, and each column is the number of reads for each ASV (svecies).

### DCNF metadata

```{r}
#| label: load-dcnf

dcnf_meta <- read_xlsx("../data/raw/DCNF_metadata.xlsx") |>
  clean_names() |>
  mutate(across(c(latitute, longitude, watertemp, conductivity,
                  chlorophyll, do, p_h, gosner_stage), as.numeric))

glimpse(dcnf_meta)
```

Explore some of this data and get basic counts

```{r}
dcnf_meta |> count(type_of_sample, family) |>arrange(desc(n))

dcnf_meta |> count(pond, genus, species) |>arrange(desc(n))

dcnf_meta |>
  filter(!is.na(gosner_stage)) |>
  summarise(min = min(gosner_stage), max = max(gosner_stage),
            mean = round(mean(gosner_stage), 1), .by = species)
```

Check some environmental variables: 
```{r}
 # 87-1 is excluded from the analysis presented in the Correa paper
dcnf_meta |>
  distinct(pond, watertemp, conductivity, chlorophyll, do, p_h, type) |>
  filter(!is.na(pond) & pond != "NA") |>
  arrange(pond)
```

```{r}
dcnf_meta |>
  distinct(pond, watertemp, conductivity, chlorophyll, do, p_h) |>
  filter(!is.na(pond) & pond != "NA") |>
  pivot_longer(-pond, names_to = "variable", values_to = "value") |>
  ggplot(aes(x = pond, y = value)) +
  geom_col(fill = "#3d6b8c") +
  facet_wrap(~variable, scales = "free_y") +
  labs(title = "Environmental variables by pond — DCNF",
       x = "Pond", y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Visual check on locations
```{r}
dcnf_meta |>
  distinct(pond, latitute, longitude, type) |>
  filter(!is.na(pond) & pond != "NA") |>
  ggplot(aes(x = longitude, y = latitute, color = type, label = pond)) +
  geom_point(size = 4) +
  ggrepel::geom_text_repel(size = 3, color = "black") +
  labs(title = "DCNF pond locations",
       x = "Longitude", y = "Latitude", color = "Fish type") +
  theme_minimal() +
  coord_fixed()
```


We want to extract the area for these ponds, so isolate the coordinates, and manually add the surface area for each of the ponds from drawing the polygons on google earth. 
```{r}
dcnf_meta |>
  distinct(pond, latitute, longitude) |>
  filter(!is.na(pond) & pond != "NA") |>
  mutate(coords = paste(latitute, longitude, sep = ", ")) |>
  mutate(area = 0)
  
```


# Match samples

I want to filter the OTU matrix to the samples from the DCNF only, 

```{r}
# all sample names in the OTU matrix
otu_samples <- rownames(otu_mat_filt)

# DCNF sample names from metadata
dcnf_samples <- dcnf_meta |>
  filter(!is.na(pond) & pond != "NA") |>
  pull(sample_name)

# check overlap
sum(otu_samples %in% dcnf_samples)
```
this is too low, so we must have an error, or maybe capitalization issues. 

```{r}
intersect(
  rownames(otu_mat),
  dcnf_meta$sample_name)
```

Sample names in the OTU matrix have been stripped of hyphens and special characters during processing (e.g. RC18-1A became RC181A), so a direct match fails. We normalize both sets of names before matching.

```{r}
# normalize: remove hyphens, uppercase
otu_samples_clean  <- toupper(gsub("-", "", otu_samples))
dcnf_samples_clean <- toupper(gsub("-", "", dcnf_samples))

cat("Matching samples:", sum(otu_samples_clean %in% dcnf_samples_clean), "\n")

# subset OTU matrix to DCNF samples
dcnf_otu_raw <- otu_mat[otu_samples_clean %in% dcnf_samples_clean, ]
dim(dcnf_otu_raw)
```

Great! now we have the 95 samples from DCNF and need to check and filter out singletons, etc. 

```{r}
dcnf_prev <- colSums(dcnf_otu_raw > 0)

table(cut(dcnf_prev, breaks = c(-1, 0, 1, 2, 5, 10, 50, Inf)))
```


Of the 23,051 ASVs in the full dataset, 17,201 are completely absent in DCNF and were detected only at the Brazil or Central Texas sites. A further 3,442 appear in only a single DCNF sample and are indistinguishable from sequencing noise. We apply a minimum prevalence threshold of 2, retaining 2,408 ASVs for downstream analysis.

```{r}
dcnf_otu_filt <- dcnf_otu_raw[, dcnf_prev >= 2]
dim(dcnf_otu_filt)
```


