---
title: "01 - DCNF sjSDM"
subtitle: "Tadpole gut microbiome: Joint Species Distribution Model — East Texas (DCNF)"
author: "Javi Rudolph"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    theme: flatly
    embed-resources: true
execute:
  warning: false
  message: false
---

## Overview

This notebook focuses on the **David Crockett and Angelina National Forests (DCNF / East Texas)**
site, which has the richest sampling in the dataset: 9 ponds, 6 tadpole species, and the most
complete set of environmental covariates (dissolved oxygen, pH, conductivity, chlorophyll-a,
water temperature, and fish presence/absence).

The goal is to fit a **Joint Species Distribution Model (sjSDM)** to the tadpole gut microbiome
at DCNF, using pond-level environmental variables as predictors and latent factors to capture
residual co-occurrence structure among ASVs after accounting for measured environment.

---

## Setup

```{r}
#| label: libraries

library(tidyverse)
library(readxl)
library(janitor)
library(sjSDM)
```

---

## Load raw data

### OTU table

The full table has 243 samples × 23,051 ASVs across all three sites (Brazil, Central Texas,
DCNF). We load it, convert to a matrix immediately, and drop the tibble to save memory.

```{r}
#| label: load-otu

otu_raw <- read_csv("../data/raw/otu_table.csv")

otu_mat <- otu_raw |>
  column_to_rownames("SampleName") |>
  as.matrix()

rm(otu_raw)
gc()

```

Since this one is heavy, we want to apply the filters we explored in the previous notebook:

```{r}
# Sparsity summary
mean(otu_mat == 0)

sample_reads <- rowSums(otu_mat)
asv_prevalence <- colSums(otu_mat > 0)

summary(sample_reads)
summary(asv_prevalence)
table(cut(asv_prevalence, breaks = c(0, 1, 2, 5, 10, 50, 100, Inf)))
```
Choose what we use as the filter: 

```{r}
# --- change this threshold ---
min_prevalence <- 2

otu_mat_filt <- otu_mat[, asv_prevalence >= min_prevalence]
```

The output matrix is filtered, where rownames are the sample name, and each column is the number of reads for each ASV (svecies).

### DCNF metadata

```{r}
#| label: load-dcnf

dcnf_meta <- read_xlsx("../data/raw/DCNF_metadata.xlsx") |>
  clean_names() |>
  mutate(across(c(latitute, longitude, watertemp, conductivity,
                  chlorophyll, do, p_h, gosner_stage), as.numeric))

glimpse(dcnf_meta)
```

Explore some of this data and get basic counts

```{r}
dcnf_meta |> count(type_of_sample, family) |>arrange(desc(n))

dcnf_meta |> count(pond, genus, species) |>arrange(desc(n))

dcnf_meta |>
  filter(!is.na(gosner_stage)) |>
  summarise(min = min(gosner_stage), max = max(gosner_stage),
            mean = round(mean(gosner_stage), 1), .by = species)
```

Check some environmental variables: 
```{r}
 # 87-1 is excluded from the analysis presented in the Correa paper
dcnf_meta |>
  distinct(pond, watertemp, conductivity, chlorophyll, do, p_h, type) |>
  filter(!is.na(pond) & pond != "NA") |>
  arrange(pond)
```

```{r}
dcnf_meta |>
  distinct(pond, watertemp, conductivity, chlorophyll, do, p_h) |>
  filter(!is.na(pond) & pond != "NA") |>
  pivot_longer(-pond, names_to = "variable", values_to = "value") |>
  ggplot(aes(x = pond, y = value)) +
  geom_col(fill = "#3d6b8c") +
  facet_wrap(~variable, scales = "free_y") +
  labs(title = "Environmental variables by pond — DCNF",
       x = "Pond", y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Visual check on locations
```{r}
dcnf_meta |>
  distinct(pond, latitute, longitude, type) |>
  filter(!is.na(pond) & pond != "NA") |>
  ggplot(aes(x = longitude, y = latitute, color = type, label = pond)) +
  geom_point(size = 4) +
  ggrepel::geom_text_repel(size = 3, color = "black") +
  labs(title = "DCNF pond locations",
       x = "Longitude", y = "Latitude", color = "Fish type") +
  theme_minimal() +
  coord_fixed()
```


We want to extract the area for these ponds, so isolate the coordinates, and manually add the surface area for each of the ponds from drawing the polygons on google earth. 
```{r}
dcnf_meta |>
  distinct(pond, latitute, longitude) |>
  filter(!is.na(pond) & pond != "NA") |>
  mutate(coords = paste(latitute, longitude, sep = ", ")) |>
  mutate(area = 0)
  
```


# Match samples

I want to filter the OTU matrix to the samples from the DCNF only, 

```{r}
# all sample names in the OTU matrix
otu_samples <- rownames(otu_mat_filt)

# DCNF sample names from metadata
dcnf_samples <- dcnf_meta |>
  filter(!is.na(pond) & pond != "NA") |>
  pull(sample_name)

# check overlap
sum(otu_samples %in% dcnf_samples)
```
this is too low, so we must have an error, or maybe capitalization issues. 

```{r}
intersect(
  rownames(otu_mat),
  dcnf_meta$sample_name)
```

Sample names in the OTU matrix have been stripped of hyphens and special characters during processing (e.g. RC18-1A became RC181A), so a direct match fails. We normalize both sets of names before matching.

```{r}
# normalize: remove hyphens, uppercase
otu_samples_clean  <- toupper(gsub("-", "", otu_samples))
dcnf_samples_clean <- toupper(gsub("-", "", dcnf_samples))

cat("Matching samples:", sum(otu_samples_clean %in% dcnf_samples_clean), "\n")

# subset OTU matrix to DCNF samples
dcnf_otu_raw <- otu_mat[otu_samples_clean %in% dcnf_samples_clean, ]
dim(dcnf_otu_raw)
```

Great! now we have the 95 samples from DCNF and need to check and filter out singletons, etc. 

```{r}
dcnf_prev <- colSums(dcnf_otu_raw > 0)

table(cut(dcnf_prev, breaks = c(-1, 0, 1, 2, 5, 10, 50, Inf)))
```


Of the 23,051 ASVs in the full dataset, 17,201 are completely absent in DCNF and were detected only at the Brazil or Central Texas sites. A further 3,442 appear in only a single DCNF sample and are indistinguishable from sequencing noise. We apply a minimum prevalence threshold of 2, retaining 2,408 ASVs for downstream analysis.

```{r}
dcnf_otu_filt <- dcnf_otu_raw[, dcnf_prev >= 2]
dim(dcnf_otu_filt)
```

# Model fitting checks

Before building the response matrix, we inspect the distribution of read counts within the DCNF subset. Microbiome count data is typically extremely right-skewed, with a few highly abundant ASVs and many low-count detections. This informs our modelling choices: presence/absence (Bernoulli/probit) sidesteps compositionality and overdispersion entirely, while a count-based model (Poisson/Negative Binomial) uses the raw integers directly with a log link. We will run both and compare. The log1p-transformed histogram gives a clearer view of the count distribution than the raw values.

```{r}
dcnf_reads <- dcnf_otu_filt[dcnf_otu_filt > 0]

summary(dcnf_reads)

hist(log1p(dcnf_reads), breaks = 100,
     main = "Distribution of log1p(reads) — DCNF non-zero counts",
     xlab = "log1p(reads)")
```

# Test fits

```{r}
# we filter to match the 95 samples we have that correspond to the ASVs data
dcnf_df <- dcnf_meta |>
  filter(!is.na(pond) & pond != "NA") |>
  filter(toupper(gsub("-", "", sample_name)) %in% 
           toupper(gsub("-", "", rownames(dcnf_otu_filt)))) |>
  arrange(match(toupper(gsub("-", "", sample_name)),
                toupper(gsub("-", "", rownames(dcnf_otu_filt)))))

# AND we add another classifier
dcnf_df <- dcnf_df |>
  mutate(
    is_tadpole  = family %in% c("Ranidae", "Hylidae", "Microhylidae"),
    is_salamander = family %in% c("Sirenidae", "Ambystomatidae", "Salamandridae"),
    is_env      = family %in% c("WATER", "SUBSTRATE")
  )

dcnf_df |> count(is_tadpole, is_salamander, is_env, family)
```

```{r}
# Y matrix presence/absence only
Y <- (dcnf_otu_filt > 0) * 1L

# check alignment that rows match rows (sample names)
all(toupper(gsub("-", "", rownames(dcnf_otu_filt))) == 
    toupper(gsub("-", "", dcnf_df$sample_name)))
```


Try a test fit, for a simple model


```{r}
X <- dcnf_df |>
  select(chlorophyll, conductivity, p_h) |>
  mutate(across(everything(), scale)) |>
  as.matrix()

model_pa <- sjSDM(
  Y        = Y,
  env      = linear(data = X, formula = ~chlorophyll + conductivity + p_h),
  biotic   = bioticStruct(df = 2L), #make it simpler because large data
  family   = binomial("probit"),
  sampling = 100L,
  se       = FALSE,
  device   = "cpu",
  iter     = 100L
)

summary(model_pa)
```
This to me says, we need to think about functional groups and how do we classify these differently? It might change if we use abundance instead of presence/absence. 
```{r}
# 1. residual co-occurrence
image(getCor(model_pa))
```

Thinking I really want to try the abundance model instead. 

```{r}
Y_counts <- dcnf_otu_filt

model_ab <- sjSDM(
  Y        = Y_counts,
  env      = linear(data = X, formula = ~chlorophyll + conductivity + p_h),
  biotic   = bioticStruct(df = 2L),
  family   = poisson("log"),
  sampling = 100L,
  se       = FALSE,
  device   = "cpu",
  iter     = 100L
)

image(getCor(model_ab))
```



```{r}
cor_mat <- getCor(model_pa)
cor_vec <- cor_mat[upper.tri(cor_mat)]

hist(cor_vec, breaks = 50,
     main = "Residual ASV co-occurrence correlations",
     xlab = "Correlation")

cat("Strong positive (r >  0.3):", sum(cor_vec >  0.3), "\n")
cat("Strong negative (r < -0.3):", sum(cor_vec < -0.3), "\n")
```


```{r}
# top 50 most prevalent ASVs

dimnames(cor_mat) <- list(colnames(dcnf_otu_filt), colnames(dcnf_otu_filt))

top50 <- names(sort(colSums(dcnf_otu_filt > 0), decreasing = TRUE)[1:20])
cor_top <- cor_mat[top50, top50]
image(cor_top)
```


We need more exploration of smaller datasets here or something. 
Next step: organize the taxa by family or something like that and visualize it like this. 


The co-occurrence matrix has 2408 × 2408 entries and is uninterpretable as a raw image. To get something meaningful, we focus on the 50 most prevalent ASVs and order them by their phylogenetic grouping using the taxonomy table. If correlated blocks align with taxonomic groups, that suggests phylogenetically related bacteria tend to co-occur, or avoid each other, in consistent ways across tadpole guts.

```{r}
tax_raw <- read_csv("../data/raw/tax_table.csv")

# get taxonomy for top 50 ASVs
tax_top50 <- tax_raw |>
  filter(SV %in% top50) |>
  arrange(Phylum, Class, Order, Family, Genus)

# reorder correlation matrix by phylogenetic order
ordered_asvs <- tax_top50$SV
cor_ordered  <- cor_mat[ordered_asvs, ordered_asvs]
```

```{r}
library(pheatmap)

# annotation for rows/cols
annot <- tax_top50 |>
  select(SV, Order) |>
  column_to_rownames("SV")

pheatmap(cor_ordered,
         annotation_row = annot,
         annotation_col = annot,
         show_rownames  = TRUE,
         show_colnames  = FALSE,
         main           = "Residual co-occurrence — top 50 ASVs ordered by taxonomy")
```


```{r}
# 3. variation partitioning
an <- anova(model_pa, verbose = FALSE)
summary(an)
plot(an)

results <- internalStructure(an)
plot(results)

```





