---
title: "01 - DCNF sjSDM"
subtitle: "Tadpole gut microbiome: Joint Species Distribution Model — East Texas (DCNF)"
author: "Javi Rudolph"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    theme: flatly
    embed-resources: true
execute:
  warning: false
  message: false
---

## Overview

This notebook focuses on the **David Crockett and Angelina National Forests (DCNF / East Texas)**
site, which has the richest sampling in the dataset: 9 ponds, 6 tadpole species, and the most
complete set of environmental covariates (dissolved oxygen, pH, conductivity, chlorophyll-a,
water temperature, and fish presence/absence).

The goal is to fit a **Joint Species Distribution Model (sjSDM)** to the tadpole gut microbiome
at DCNF, using pond-level environmental variables as predictors and latent factors to capture
residual co-occurrence structure among ASVs after accounting for measured environment.

---

## Setup

```{r}
#| label: libraries

library(tidyverse)
library(readxl)
library(janitor)
library(sjSDM)
```

---

## Load raw data

### OTU table

The full table has 243 samples × 23,051 ASVs across all three sites (Brazil, Central Texas,
DCNF). We load it, convert to a matrix immediately, and drop the tibble to save memory.

```{r}
#| label: load-otu

otu_raw <- read_csv("../data/raw/otu_table.csv")

otu_mat <- otu_raw |>
  column_to_rownames("SampleName") |>
  as.matrix()

rm(otu_raw)
gc()

```

### DCNF metadata

```{r}
#| label: load-dcnf

dcnf_meta <- read_xlsx("../data/raw/DCNF_metadata.xlsx") |>
  clean_names() |>
  mutate(across(c(latitute, longitude, watertemp, conductivity,
                  chlorophyll, do, p_h, gosner_stage), as.numeric))

glimpse(dcnf_meta)
```

Explore some of this data and get basic counts

```{r}
dcnf_meta |> count(type_of_sample, family) |>arrange(desc(n))

dcnf_meta |> count(pond, genus, species) |>arrange(desc(n))

dcnf_meta |>
  filter(!is.na(gosner_stage)) |>
  summarise(min = min(gosner_stage), max = max(gosner_stage),
            mean = round(mean(gosner_stage), 1), .by = species)
```

Check some environmental variables: 
```{r}
 # 87-1 is excluded from the analysis presented in the Correa paper
dcnf_meta |>
  distinct(pond, watertemp, conductivity, chlorophyll, do, p_h, type) |>
  filter(!is.na(pond) & pond != "NA") |>
  arrange(pond)
```

```{r}
dcnf_meta |>
  distinct(pond, watertemp, conductivity, chlorophyll, do, p_h) |>
  filter(!is.na(pond) & pond != "NA") |>
  pivot_longer(-pond, names_to = "variable", values_to = "value") |>
  ggplot(aes(x = pond, y = value)) +
  geom_col(fill = "#3d6b8c") +
  facet_wrap(~variable, scales = "free_y") +
  labs(title = "Environmental variables by pond — DCNF",
       x = "Pond", y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Visual check on locations
```{r}
dcnf_meta |>
  distinct(pond, latitute, longitude, type) |>
  filter(!is.na(pond) & pond != "NA") |>
  ggplot(aes(x = longitude, y = latitute, color = type, label = pond)) +
  geom_point(size = 4) +
  ggrepel::geom_text_repel(size = 3, color = "black") +
  labs(title = "DCNF pond locations",
       x = "Longitude", y = "Latitude", color = "Fish type") +
  theme_minimal() +
  coord_fixed()
```


We want to extract the area for these ponds, so isolate the coordinates, and manually add the surface area for each of the ponds from drawing the polygons on google earth. 
```{r}
dcnf_meta |>
  distinct(pond, latitute, longitude) |>
  filter(!is.na(pond) & pond != "NA") |>
  mutate(coords = paste(latitute, longitude, sep = ", ")) |>
  mutate |>
  
```
